Bayesian Rate Estimation

### 2012-08-16 many.tex
Some edits to cluster example.

Our final example concerns fitting for the location and shape
parameters of a cluster of stars observed on top of a stellar
background with a density gradient.  We assume that a star cluster has
a Plummer surface-density profile \citep{Plummer1911,Aarseth1974},
\begin{equation}
  \label{eq:plummer-surface-density}
  \hat{f}(\vec{x}, \theta) = \frac{1}{\pi r_0^2 \left( 1 +
    \frac{\left| \vec{x} - \vec{x}_0 \right|^2}{r_0^2} \right)^2},
\end{equation}
where $\vec{x}_0$ is the location on the sky of the center of the
cluster, and $r_0$ is a radial scale parameter.  We assume a square
observational domain\footnote{The observational domain is not
  infinite, so the normalization of the cluster density in
  Eq.~\eqref{eq:plummer-surface-density} is not quite correct.  In our
  modeling we properly take this into account, but for simplicity here
  we ignore it.}, $\vec{x} \in [0,1]^2$, and a background that has a
density gradient at an arbitrary orientation with respect to the
observational axes:
\begin{equation}
  \hat{b}\left(\vec{x}, \theta\right) = 1 + \vec{\gamma} \cdot \left(
  \vec{x} - \vec{x}_{1/2} \right),
\end{equation}
where $\vec{\gamma}$ is the gradient, and $\vec{x}_{1/2} = [1/2, 1/2]$
is the centroid of the observational domain.  \ilya{[I think that if
    the domain is finite, the previous equation is only normalized for
    all $\vec{\gamma}$ if $\vec{x}_{1/2}$ is in the center of a
    symmetric domain (as it happens to be in this case).]}  We use
assume a gradient for the background of $\vec{\gamma} = [-0.5, 0.5]$.
\ilya{[I find it a bit confusing to use the word "assume" for both the
    model and for the specific parameters injected into the simulation
    (but then being fit for, so their values are not "assumed" in the
    recovery).  Maybe "assume" for the model, and "simulate" for the
    parameter values to be fit?]}  We assume that the total number of
background stars, $R_b = 10000$, exceeds by a factor of ten the total
number of cluster stars, $R_f = 1000$.  \ilya{[But the cluster has
    only $\sim 0.01$ of the total area, so it's over-dense by a factor
    of $10$ or more.  Fig.~6 shows a very bright spot where the
    cluster is, which sort of makes the problem look too simple for
    the chosen parameters...]}  We choose a true cluster center
$\vec{x}_0 = \vec{x}_{1/2}$, and a radial scale parameter of $r_0 =
0.1$.  In all, the shape parameters for the true distributions are
\begin{equation}
\label{eq:true-cluster-parameters}
\theta_0 \equiv \left\{ x_0, y_0, r_0, \gamma_x, \gamma_y \right\} =
\left\{ \frac{1}{2}, \frac{1}{2}, \frac{1}{10}, -\frac{1}{2},
\frac{1}{2} \right\}.
\end{equation}
Figure \ref{fig:sky-density} shows the density on the sky due to the
cluster foreground and the background with these parameters.  We drew
a synthetic data set from this density according to the corresponding
inhomogeneous Poisson process.

\begin{figure}
  \includegraphics[width=\columnwidth]{sky-density}
  \caption{\label{fig:sky-density} The density contours on the sky of
    the forergound cluster and background discussed in \S
    \ref{sec:star-cluster}, with true parameters given in
    Eq.~\eqref{eq:true-cluster-parameters}.  There are a factor of 10
    more stars (10000) in the background than in the cluster (1000),
    but the peak density for the cluster is much higher.  \ilya{[It
        seems too obvious that where the cluster is by eye, and
        probably fairly straightforward to compute the number of
        "excess" stars there...  Reader might ask why we need
        something so complicated when the answer is seems so obvious
        and a "frequentist" approach would seem to do here.]}}
\end{figure}

To analyze our synthetic data set, we analytically marginalized over
the state flags, using the likelihood in
Eq.~\eqref{eq:rate-shape-posterior}.  \ilya{[I think it would be kind
    to the reader to be more explicit about the connection with the
    formalism in Section II.  What are the "events"?  What is the
    ranking statistic $x$?  What are the flags (which aren't in
    $\mathbb{R}$?)?]}  We did this to take advantage of the
\textit{emcee} sampler of \citet{ForemanMackey2012}, which requires
all parameters to be in $\mathbb{R}$.  We applied a prior on the shape
parameters that is flat in $\vec{x}_0$ and $\vec{\gamma}$, and the
Jeffreys prior on $r_0$,
\begin{equation}
  p\left( r_0 \right) = \frac{\sqrt{R_f}}{r_0}.
\end{equation}
\ilya{[Why is this the Jeffreys prior (BTW, it's Jeffreys, or perhaps
    Jeffreys', but definitely not Jeffrey's ;) )?  It's not obvious
    why the prior on $r_0$ should depend on $R_f$?]}  (Note that this
factor of $\sqrt{R_f}$ cancels with the Jeffreys prior on the rate,
$1/\sqrt{R_f}$; we have verified that the priors on these parameters
are irrelevant to our results, as would be expected from the
measurement of $\sim 1000$ foreground stars.)

Figure \ref{fig:sky-loc} shows the posterior for the location
parameters, $\vec{x}_0$; the center of the cluster is localized to
within a few percent of its scale.  Figure \ref{fig:cluster-number}
shows the posteriors inferred on the cluster and background numbers,
$R_f$ and $R_b$, and Figure \ref{fig:cluster-scale} shows the
posterior for the cluster's scale parameter.  In spite of the
significant background, the cluster scale and total number are
accurately recovered by our analysis.

>>>>

\will{Look up von Hippel paper about Bayes 9, the cluster isochrone
  fitting software---we should cite, since our approach to cluster
  membership.}

Our final example concerns fitting for the location and shape
parameters of a cluster of stars observed on top of a stellar
background with a density gradient.  In this example, stars are either
members of the cluster (i.e.~foreground) or background contamination,
with a spatially varying density (i.e.~our rate functions are
two-dimensional).  We assume that a star cluster has a Plummer
surface-density profile \citep{Plummer1911,Aarseth1974},
\begin{equation}
  \label{eq:plummer-surface-density}
  \hat{f}(\vec{x}, \theta) = \frac{1}{\pi r_0^2 \left( 1 +
    \frac{\left| \vec{x} - \vec{x}_0 \right|^2}{r_0^2} \right)^2},
\end{equation}
where $\vec{x}_0$ is the location on the sky of the center of the
cluster, $r_0$ is a radial scale parameter, and $\vec{x} = \left( x, y
\right)$ is the position on the sky.  We assume a square observational
domain\footnote{The observational domain is not infinite, so the
  normalization of the cluster density in
  Eq.~\eqref{eq:plummer-surface-density} is not quite correct.  In our
  modeling we properly take this into account, but for simplicity here
  we ignore it.}, $\vec{x} \in [0,1]^2$, and a background that has a
density gradient at an arbitrary orientation with respect to the
observational axes:
\begin{equation}
  \hat{b}\left(\vec{x}, \theta\right) = 1 + \vec{\gamma} \cdot \left(
  \vec{x} - \vec{x}_{1/2} \right),
\end{equation}
where $\vec{\gamma}$ is the gradient, and $\vec{x}_{1/2} = [1/2, 1/2]$
is the centroid of the observational domain.  \ilya{[I think that if
    the domain is finite, the previous equation is only normalized for
    all $\vec{\gamma}$ if $\vec{x}_{1/2}$ is in the center of a
    symmetric domain (as it happens to be in this case).]}  \will{Yes,
  as you say, this function is only normalized for $\vec{x}_{1/2}$ is
  the center, and also for small enough $\vec{\gamma}$; both of these
  conditions obtain here.}

We use simulated data drawn from our model with parameters
\begin{equation}
\label{eq:true-cluster-parameters}
\theta_0 \equiv \left\{ x_0, y_0, r_0, \gamma_x, \gamma_y \right\} =
\left\{ \frac{1}{2}, \frac{1}{2}, 0.18, -\frac{1}{2},
\frac{1}{2} \right\},
\end{equation}
with $R_f = 1000$ and $R_b = 10000$.  For this set of parameters, the
average density of the background and the peak density of the cluster
are comparable; there are an order of magnitude more background stars
than cluster stars in the field.  Figure \ref{fig:sky-density} shows
the density of stars (cluster and background) on the sky for our
particular data set.

\begin{figure}
  \includegraphics[width=\columnwidth]{sky-density}
  \caption{\label{fig:sky-density} The density contours on the sky of
    our synthetic data set of cluster and background stars for the
    parametrs in Eq.~\eqref{eq:true-cluster-parameters}.  There are a
    factor of 10 more stars (10000) in the background than in the
    cluster (1000), but the average density of cluster and background
    stars is similar.}
\end{figure}

To analyze our synthetic data set, we analytically marginalized over
the state flags (i.e.~cluster membership), using the likelihood in
Eq.~\eqref{eq:rate-shape-posterior}.  We did this to take advantage of
the \textit{emcee} sampler of \citet{ForemanMackey2012}, which
requires all parameters to be in $\mathbb{R}$.  We applied a prior on
the shape parameters that is flat in $\vec{x}_0$ and $\vec{\gamma}$,
and an (approximately) Jeffreys prior on $r_0$,
\begin{equation}
  p\left( r_0 \right) = \frac{\sqrt{R_f}}{r_0}.
\end{equation}
\ilya{[Why is this the Jeffreys prior (BTW, it's Jeffreys, or perhaps
    Jeffreys', but definitely not Jeffrey's ;) )?  It's not obvious
    why the prior on $r_0$ should depend on $R_f$?]}  \will{Here is
  the calculation:
\begin{equation}
  p\left( r_0 \right) = \sqrt{\left\langle \left(\frac{\partial \log
      \mathcal{L}}{\partial r_0 } \right)^2 \right\rangle} = \sqrt{
    R_f \left\langle \left(\frac{\partial \log\hat{f}}{\partial r_0
    }\right)^2 \right\rangle } = \sqrt{R_f \left\langle \left(
    \frac{2}{r_0} - \frac{4 r_0}{r^2 + r_0^2} \right)^2 \right\rangle}
  = \sqrt{\frac{ 4 R_f}{3 r_0^2}} \propto \frac{\sqrt{R_f}}{r_0}.
\end{equation}
Of course, we are ignoring the fact that $R_f$ and $r_0$ are
correlated.  In truth, the Jeffreys prior would be the determinant of
the matrix built out of terms like $\partial \log\mathcal{L} /
\partial \theta_i \times \partial \log\mathcal{L} /\partial \theta_j$,
but let's not get carried away\ldots.  As another aside, note that we
could have predicted the $1/r_0$ dependence because $r_0$ has units,
so the only unit-invariant prior is to be flat in $\log r_0$; I'm
don't have a similar argument for the appearance of the $\sqrt{R_f}$
factor.}  (Note that this factor of $\sqrt{R_f}$ cancels with the
Jeffreys prior on the rate, $1/\sqrt{R_f}$; we have verified that the
priors on these parameters are irrelevant to our results, as would be
expected from the measurement of $\sim 1000$ foreground stars.)

Figure \ref{fig:sky-loc} shows the posterior for the location
parameters, $\vec{x}_0$; the center of the cluster is localized to
within about 10\% of the cluster scale.  Figure
\ref{fig:cluster-number} shows the posteriors inferred on the cluster
and background numbers, $R_f$ and $R_b$, and Figure
\ref{fig:cluster-scale} shows the posterior for the cluster's scale
parameter.  In spite of the significant background, the cluster scale
and total number are accurately recovered by our analysis.
